---
title: "Calculating effect sizes"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(gdata,quietly=TRUE)
require(ggplot2)
require(dplyr)
require(metafor)
```

## Load packages

```
require(gdata)
require(metafor)
require(dplyr)
require(ggplot2)

```


## Download data

For this exercise, we'll use the Curtis et al. (1999) data set that contains an array of plant measurements taken on 65 tree species under control and elevated CO~2~ levels 
This data has been extracted from 84 papers and includes mean, standard deviation, coefficient of variation, and 
number of observations.  



```{r curtis,echo=TRUE}
curtis<-read.xls("http://www.nceas.ucsb.edu/meta/Curtis/Curtis_CO2_database.xls",as.is=TRUE,verbose=FALSE,sheet=1)

str(curtis)
```

## Calculate effect sizes 

For the Curtis data set, the response variables are all measured on a continuous scale. This allows us to calculate a number of
traditional effect sizes, including Hedge's D and the log-response ratio. In 'metafor' these effect sizes are referred to as 'SMD' and 'ROM', respectively.  

In this case, we need the following variables to calculate effect sizes: mean, standard deviation, and sample size. 

Additionally, sampling variances can be estimated differently. The default setting (used here) is the 'large-sample approximation' (vtype='LS'). For Hedge's D,
one can also estimate 'unbiased estimates' (vtype='UB'); for log-response ratio, 'LS' calculates sampling variance without assuming homoscedasticity while 'HO' assumes that variances are the same in control and treatment groups.  

Note that 'group 1' (e.g. 'm1i' , 'sd1i') correspond to the treatment group and 'group 2' (e.g. 'm2i', 'sd2i') correspond to the control group.  

### Hedges' D: Photosynthesis  

```{r curtis_ES, echo=TRUE}

curtis_ES<-escalc(measure='SMD', m2i=X_AMB , sd2i=SD_AMB, n2i=N_AMB, m1i=X_ELEV, sd1i=SD_ELEV, n1i=N_ELEV, vtype='LS',var.names=c("Hedges_D","Hedges_var"),data=curtis)

```


## Visualize effect sizes (part I): forest plots     

Please note that the first 25 observations are presented for visualization reasons.  If you want to look at all effect sizes,  
use the name of the larger data frame ('hedges_PN')  in the 'forest' function.  

```{r forestplot, echo=TRUE}

hedges_PN<-filter(curtis_ES, PARAM=="PN")
hedges_PN<-arrange(hedges_PN, GENUS)
hedges_PNN<-hedges_PN[1:25,]  # not necessary

forest(hedges_PNN$Hedges_D,hedges_PNN$Hedges_var, slab=hedges_PNN$GENUS, showweights=FALSE)

```

__QUESTION: Do trees have higher or lower photosynthesis under ambient or elevated CO~2~ conditions?__



## Visualize effect sizes (part II): explore moderators    

Alternatives for visually assessing effect sizes (per group) include histograms or density distributions.  

```{r forestplott, echo=TRUE}

hedges_PN$SE<-sqrt(hedges_PN$Hedges_var)

dodge <- position_dodge(width=1)

ggplot(hedges_PN, aes(x=DIV2, y=Hedges_D, colour=DIV2)) + 
    geom_hline(yintercept=0,color="red")+
    #geom_errorbar(aes(ymin=Hedges_D-SE, ymax=Hedges_D+SE), width=.1,position=dodge) +
    geom_point(position=dodge)+ labs(x="Plant group", y="Effect size")+
     guides(fill=FALSE,colour=guide_legend(title="Plant group",title.position = "top"))+
  
  theme_bw()




```

__QUESTION: Are there differences among functional groups in terms of their response to elevated CO~2~?#__



### LRR: water-use efficiency  

The log-response ratio is a commonly used effect size and -- once back-transformed -- is easier to understand (e.g. the mean effect 
size is 10% greater than the control).  

```{r curtis_ES2, echo=TRUE}
require(metafor,quietly = TRUE)

curtis_ES2<-escalc(measure='ROM', m2i=X_AMB , sd2i=SD_AMB, n2i=N_AMB, m1i=X_ELEV, sd1i=SD_ELEV, n1i=N_ELEV, vtype='LS',var.names=c("LRR","LRR_var"),data=curtis)

str(curtis_ES2)

```


## Visualize effect sizes (part I) : forest plots     

```{r forestplot2, echo=TRUE}
require(metafor,quietly = TRUE)
require(dplyr,quietly=TRUE)

LRR_WUE<-filter(curtis_ES2, PARAM=="WUE")
LRR_WUE<-arrange(LRR_WUE, GENUS)

forest(LRR_WUE$LRR,LRR_WUE$LRR_var, slab=LRR_WUE$GENUS, showweights=TRUE)
```

Note that the percentage next to each effect size is the weight of each point.  


__QUESTION: Do trees have higher or lower water-use efficiency in response to elevated CO~2~?__
