---
title: "Hierarchical meta-analytical models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(gdata,quietly=TRUE)
require(metafor)
require(dplyr)
require(multcomp)
require(ggplot2)
```

## Getting started   

_Load packages_ 

```
require(gdata)
require(metafor)
require(dplyr)
require(multcomp)
require(ggplot2)

```

_Download data (Curtis et al. 1999)_    


```{r dataa, echo=TRUE}

curtis<-read.xls("http://www.nceas.ucsb.edu/meta/Curtis/Curtis_CO2_database.xls",as.is=TRUE,verbose=FALSE,sheet=1)
curtis_ES<-escalc(measure='ROM', m2i=X_AMB , sd2i=SD_AMB, n2i=N_AMB, m1i=X_ELEV, sd1i=SD_ELEV, n1i=N_ELEV, vtype='LS',var.names=c("LRR","LRR_var"),data=curtis)
summary(as.factor(curtis_ES$PARAM))
curtis_WT<-filter(curtis_ES, PARAM=="TOTWT") # let's use whole plant weight because it has the largest number of observations   

curtis_WT$GEN_SPP<-paste(curtis_WT$GENUS,curtis_WT$SPECIES,sep="_")
```


## Part I: compare fixed and random effects models  

### Fixed effects model

**Important** Fixed effects models assume that there is one true effect size, i.e. differences are largely due to sampling error.      

```{r FEmod, echo=TRUE}

fix_wt<-rma(LRR, LRR_var, method="FE", data=curtis_WT)

summary(fix_wt)
```

This model estimates the 'grand mean' of the effect of CO~2~ exposure on total plant weight. 



### Random effects model

**Important** Random effects models allows the true effect size to differ.  
Here, the effect sizes represent a random sample from a particular distribution.   

You can use either the 'rma' function, which uses REML as the default method for fitting a model, or 'rma.mv', which
requires that you specify the random effects.  

Here, we use a random group term for each study ('PAP_NO') because many studies report more than one effect size.

```{r obz , echo=FALSE}
a<-summarize(group_by(curtis_WT, PAP_NO), obs=length(unique(OBSNO)))
mean(a$obs) # mean number of observations per study  
```



```{r REmod, echo=TRUE}

re_wt<-rma.mv(LRR, LRR_var, random=~1|PAP_NO, data=curtis_WT)

summary(re_wt)
```
This model also estimates the 'grand mean' of the effect of CO~2~ exposure on total plant weight.  


```{r comps, echo=FALSE}
fix<-cbind("model"="FE","estimate"=fix_wt$b,"lCI"=fix_wt$ci.lb,hCI=fix_wt$ci.ub)

random<-cbind("model"="RE","estimate"=re_wt$b,"lCI"=re_wt$ci.lb,hCI=re_wt$ci.ub)

togg<-rbind.data.frame(fix,random)
colnames(togg)[2]<-"Estimate"

togg$Estimate<-as.numeric(as.character(togg$Estimate))
togg$lCI<-as.numeric(as.character(togg$lCI))
togg$hCI<-as.numeric(as.character(togg$hCI))


plott<-ggplot(togg, aes(x=model, y=Estimate, colour=model)) + 
    geom_hline(yintercept=0,color="red")+
    geom_errorbar(aes(ymin=lCI, ymax=hCI), width=.1) +
    geom_point(size=1)+ labs(x="Model", y="Effect size")+
     guides(fill=FALSE,colour=guide_legend(title="Model type",title.position = "top"))+
  
  theme_bw()+theme(axis.text.x=element_text(face="bold"),axis.text.y=element_text(face="bold"))
plott
```
 
__Comparison of model heterogeneity__


Note that _I^2^_ for the random effects model integrates heterogeneity from within- and between-clusters following Nakagawa & Santos (2012).  

See [here](http://www.metafor-project.org/doku.php/tips:i2_multilevel_multivariate) to calculate _I^2^_ for between- and within-clusters of multilevel models separately.  
```{r i2, echo=TRUE}

I2_fe<-cbind.data.frame("model"="fixed","I2"=fix_wt$I2) 

#######################################
# calcualte I2 for a multilevel model #
#######################################
W <- diag(1/curtis_WT$LRR_var)
X <- model.matrix(re_wt)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
re_I2<-100 * sum(re_wt$sigma2) / (sum(re_wt$sigma2) + (re_wt$k-re_wt$p)/sum(diag(P)))

I2_re<-cbind.data.frame("model"="random","I2"=re_I2)

#######################################

I22<-rbind.data.frame(I2_fe,I2_re)
I22
```



## Part II: Hierarchical, multi-level model ('meta-regression')  


__Identify most parsimonious random-effects structure__  

I've included additional random effects included in the data set: extra treatment (e.g. nutrient addition, light, water), species identity, and 
a combination of the two.  



```{r REmod_re, echo=TRUE}

re_wt1<-rma.mv(LRR, LRR_var, mods=~DIV2,random=~1|PAP_NO, data=curtis_WT)

re_wt2<-rma.mv(LRR, LRR_var, mods=~DIV2,random=list(~1|PAP_NO, ~1|XTRT), data=curtis_WT)

re_wt3<-rma.mv(LRR, LRR_var, mods=~DIV2,random=list(~1|PAP_NO, ~1|GEN_SPP), data=curtis_WT)

re_wt4<-rma.mv(LRR, LRR_var, mods=~DIV2,random=list(~1|PAP_NO, ~1|XTRT, ~1|GEN_SPP), data=curtis_WT)

AICc<-rbind(mod1=re_wt1$fit.stats$REML[5],mod2=re_wt2$fit.stats$REML[5],mod3=re_wt3$fit.stats$REML[5],mod4=re_wt4$fit.stats$REML[5])

AICc  # and the winner is ... 

```

__Test significance of moderators__  

There are a couple of ways to test the signifance of moderator variables.  

_Likelihood Ratio Tests_  

Note that maximum likelihood is used to fit both models  

```{r sigg, echo=TRUE}

bigg<-rma.mv(LRR, LRR_var, mods=~DIV2,random=list(~1|PAP_NO, ~1|XTRT, ~1|GEN_SPP), data=curtis_WT, method="ML")
small<-rma.mv(LRR, LRR_var, mods=~1,random=list(~1|PAP_NO, ~1|XTRT, ~1|GEN_SPP), data=curtis_WT, method="ML")

anova(bigg,small)
```

_Knapp \& Hartung Adjustment_  

The Knapp \& Hartung adjustment, similar to the likelihood ratio test, can be used for individual model coefficients (t-tests) or a group of model coefficients (F tests).  



```{r blaa, echo=TRUE}

bigg<-rma.mv(LRR, LRR_var, mods=~DIV2,random=list(~1|PAP_NO, ~1|XTRT, ~1|GEN_SPP), data=curtis_WT, test="knha")

summary(bigg)

```

In both cases, including plant group as a moderator variable did not add information to the model.  


_Model Heterogeneity (I^2^)_ 


```{r I22, echo=TRUE}

W <- diag(1/curtis_WT$LRR_var)
X <- model.matrix(bigg)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
100 * sum(bigg$sigma2) / (sum(bigg$sigma2) + (bigg$k-bigg$p)/sum(diag(P)))

```

Note that the _I^2^_ for this model is very similar to that of the fixed effects model and the random effects model with a simpler random effects structure (seen earlier).  

__Pairwise comparisons__

We will use the 'multcomp' package to test contrasts between levels of the factor 'DIV2'.    

```{r pairz, echo=TRUE}
require(multcomp)

bigg_int<-rma.mv(LRR, LRR_var, mods=~DIV2-1,random=list(~1|PAP_NO, ~1|XTRT, ~1|GEN_SPP), data=curtis_WT, test="knha")

summary(bigg_int)
summary(glht(bigg_int, linfct=rbind(c(-1,1,0), c(-1,0,1), c(0,-1,1))), test=adjusted("none"))
```


```{r compz, echo=FALSE}


out<-cbind.data.frame("groups"=rownames(bigg_int$b),"estimate"=bigg_int$b,"lCI"=bigg_int$ci.lb,hCI=bigg_int$ci.ub)

colnames(out)[2]<-"Estimate"

out$groups<-as.character(out$groups)
out$groups<-ifelse(out$groups=="DIV2ANGIO","Angiosperm",out$groups)
out$groups<-ifelse(out$groups=="DIV2GYMNO","Gymnosperm",out$groups)
out$groups<-ifelse(out$groups=="DIV2N2FIX","N-fixers",out$groups)

plott2<-ggplot(out, aes(x=groups, y=Estimate, colour=groups)) + 
    geom_hline(yintercept=0,color="red")+
    geom_errorbar(aes(ymin=lCI, ymax=hCI), width=.1) +
    geom_point(size=1)+ labs(x="Plant groups", y="Effect size (LRR)")+
     guides(fill=FALSE,colour=guide_legend(title="Plant groups",title.position = "top"))+
  
  theme_bw()+theme(axis.text.x=element_text(face="bold"),axis.text.y=element_text(face="bold"))
plott2
```


## Exercise

1) download data from our Github repository  

```
install.packages('RCurl')

require(RCurl)
landuse <- getURL("https://raw.githubusercontent.com/dylancraven/MetaAnalysis_Course/gh-pages/Slides/LandUseBiodiv.csv")
landuse <- read.csv(text = landuse,header=T)
```

2) Fit a multi-level model

3) Calculate I^2^  

4) Test for multiple comparisons (if using a categorical moderator)  






