---
title: "Confounding effects and extra tricks"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(gdata,quietly=TRUE)
require(metafor)
require(dplyr)
#require(multcomp)
require(ggplot2)
require(compute.es,quietly=TRUE)
require(cowplot)
require(pez,quietly=TRUE)
require(phytools,quietly=TRUE)
require(ape)
setwd('/homes/dc78cahe/Dropbox (iDiv)/Teaching/MetaAnalysis_Course/pages/Day4_files/')
```


## Getting started   

We'll continue using the same data from previous days.  

_Load packages_ 

```
require(gdata)
require(metafor)
require(dplyr)
require(compute.es)
require(ggplot2)
require(cowplot)
require(pez)
require(phytools)
require(ape)
```

_Download data (Curtis et al. 1999)_    


```{r dataa, echo=TRUE}

curtis<-read.xls("http://www.nceas.ucsb.edu/meta/Curtis/Curtis_CO2_database.xls",as.is=TRUE,verbose=FALSE,sheet=1)
curtis_ES<-escalc(measure='ROM', m2i=X_AMB , sd2i=SD_AMB, n2i=N_AMB, m1i=X_ELEV, sd1i=SD_ELEV, n1i=N_ELEV, vtype='LS',var.names=c("LRR","LRR_var"),data=curtis)
#summary(as.factor(curtis_ES$PARAM))
curtis_WT<-filter(curtis_ES, PARAM=="TOTWT") # let's use whole plant weight because it has the largest number of observations   

curtis_WT$GEN_SPP<-paste(curtis_WT$GENUS,curtis_WT$SPECIES,sep="_")
```


## Conversion among effect sizes  

'Compute.es' is a powerful package that converts effect sizes.

The main function for calculating effect sizes is _mes_ and  _des_ ,and _res_ convert among effect sizes.  

Here, we will first calculate basic effect sizes using _mes_ and then convert the effect size _r_ to Fisher's _z_


```{r hedgez, echo=TRUE}
# calculate effect sizes
curtis_ES<-mes(m.2=curtis_WT$X_AMB, m.1=curtis_WT$X_ELEV, sd.2=curtis_WT$SD_AMB, sd.1=curtis_WT$SD_ELEV, n.2=curtis_WT$N_AMB, n.1=curtis_WT$N_ELEV ,verbose=FALSE)


# convert correlation coefficient to fisher's z  
new_ES<-res(r=curtis_ES$r,var.r=curtis_ES$var.r, n=curtis_ES$N.total,verbose=FALSE)

curtis_ESS<-dplyr::select(new_ES, r, var.r, N.total, fisher.z, var.z)

head(curtis_ESS)

r_z<-ggplot(curtis_ESS, aes(x=r, y=fisher.z))+geom_point() +
  
    geom_abline(intercept = 0, slope = 1,colour="red")+theme_bw()

r_z
```


```{r r_fish, echo=TRUE}

hist_r<-ggplot(curtis_ESS, aes(r)) +
  geom_density(colour="red")+xlab("r")+theme_bw()


hist_z<-ggplot(curtis_ESS, aes(fisher.z)) +
  geom_density(color="blue")+xlab("Fisher's Z")+theme_bw()

ab<-plot_grid(hist_r, hist_z, ncol=2)
ab
```


## Cumulative meta-analysis  

This method tests whether effect sizes have shifted over time. It fits the model by iteratively adding observations in the order that we designate.  


```{r model, echo=TRUE}

re_wt<-rma(LRR, LRR_var, data=curtis_WT)

cum_re<-cumul(re_wt, order(curtis_WT$OBSNO))

forest.cumul.rma(cum_re)
```


## Controlling for shared evolutionary history (phylogeny)  

For species-level analyses, shared evolutionary history should be controlled for above and beyond random effect terms already included.  

__Clean data set__

The basic steps are:  

1) Resolve taxonomic names to ensure that as many as possible can be placed on a phylogeny.  
2) Build phylogeny. If you're interested in how to build a phylogeny, here's the [code](https://dylancraven.github.io/MetaAnalysis_Course/pages/Day4_extra.html)   
3) Make sure that your phylogeny has the same number of tip labels as there are species in the data set.  



```{r merge, echo=TRUE}
setwd('/homes/dc78cahe/Dropbox (iDiv)/Teaching/MetaAnalysis_Course/pages/Day4_files/')
clean<-read.csv("TPL_sppnames.csv")
clean<-dplyr::select(clean,GEN_SPP2=Taxon, phy=new_species)
clean$GEN_SPP2<-as.character(clean$GEN_SPP2)

curtis_WT$GENUS<-tolower(as.character(curtis_WT$GENUS))
curtis_WT$GENUS<-paste(toupper(substr(curtis_WT$GENUS, 1, 1)), substr(curtis_WT$GENUS, 2, nchar(curtis_WT$GENUS)), sep="")

curtis_WT$SPECIES<-tolower(as.character(curtis_WT$SPECIES))
curtis_WT$GEN_SPP2<-as.character(paste(curtis_WT$GENUS, curtis_WT$SPECIES,sep=" "))

curtis_WT$GEN_SPP2<-ifelse(curtis_WT$GEN_SPP2=="Populusx euramericana","Populus × euramericana",curtis_WT$GEN_SPP2)

curtis_WT<-dplyr::left_join(curtis_WT,clean, by="GEN_SPP2")

# read in tree

tree<-read.tree("Curtis_phylogeny.tre")
str(tree)

# we need to drop one species from our data frame ('Trichospermum mexicanum' because it wasn't placed on the phylogeny)

curtis_WTT<-filter(curtis_WT, phy!="Trichospermum_mexicanum")

length(unique(curtis_WTT$phy))

#same number of species on phylogeny as in data set?

length(unique(curtis_WTT$phy))==length(unique(tree$tip.label))

```

__Fit multi-level meta-analytical model that accounts for shared phylogenetic history__ 


1) Make a phylogenetic correlation matrix  
2) Fit model such that there is a random term for species, which is then matched to a correlation matrix ('R')
3) Compare it to a model that doesn't account for phylogeny  

```{r phylo2, echo=TRUE}

tree_m<-vcv.phylo(tree, cor=TRUE) # creates phylogenetic correlation matrix

re_phy<-rma.mv(LRR, LRR_var, mods=~1,random=list(~1|PAP_NO, ~1|XTRT, ~1|phy), R=list(phy=tree_m), data=curtis_WTT)
summary(re_phy)

#
re_nophy<-rma.mv(LRR, LRR_var, mods=~1,random=list(~1|PAP_NO, ~1|XTRT, ~1|phy), data=curtis_WTT)
summary(re_nophy)
```

Accountring for phylogeny, in this particular case, did not alter the mean effect size. However, the confidence intervals around the mean effect size are wider when accounting for phylogeny.  


### Exercise: do your own meta-analysis (using either Gibson et al. or another data set of your own choosing)   


#Tasks#  

1. Download data from [Gibson et al.](http://www.nature.com/nature/journal/v478/n7369/abs/nature10425.html#supplementary-information)  
2. Calculate effect sizes  
2. Fit multi-levelr andom effects model  
3. Make visualisations of your results (effect sizes, forest plot, etc.)  
